\section{Construction of Free Commutative Monoids}
\label{sec:commutative-monoids}

THe next step for us is to add commutativity -- extending our constructions of free monoids to free commutative monoids.
%
Informally, adding commutativity to free monoids turns ``ordered lists'' to ``unordered lists'',
where the ordering is the one imposed by the position or index of the elements in the list.
%
This is crucial to our goal of studying sorting,
as we will study sorting as a function from unordered lists to ordered lists.
which is later in~\cref{sec:sorting}.

It is known that finite multisets are commutative monoids (in fact, free commutative monoids),
under the operation of multiset union: $xs \cup ys = ys \cup xs$.
%
Its order is ``forgotten'' in the sense that it doesn't matter how two multisets are union-ed together,
or in more concrete terms, for example, $\{a, a, b, c\} = \{b, a, c, a\}$ are equal as finite multisets
(hence justifying the set notation).
%
This is unlike free monoids, using $\List$ for example,
where $[a, a, b, c] \neq [b, a, c, a]$ (hence justifying the list notation).

\subsection{Free monoids with a quotient}\label{cmon:qfreemon}

Instead of constructing free commutative monoids directly, the first construction we study is to take a free monoid and
impose commutativity on it.
%
We do this generally, by giving a construction of free commutative monoid as a quotient of \emph{any} free monoid.
%
Specific instances of this construction are given in \cref{cmon:plist} and \cref{cmon:bag}.

From the universal algebraic perspective developed in~\cref{sec:universal-algebra},
we can ask what it means to extend the equational theory of a given algebraic signature --
or how to construct a free commutative monoid as a quotient of a free monoid.
%
If $(\str{F}(A), \eta)$ is a free monoid construction satisfying its universal property,
then we'd like to quotient $F(A)$ by an \emph{appropriate relation} $\approx$,
that turns it into a free commutative monoid.
%
This is exactly the specification of a \emph{permutation relation}!

% $(\str{F}(A), i, \mult)$ quotiented by a permutation relation $({F(A) / \approx}, e, \doubleplus)$
% would be the free commutative monoid on $A$.

\begin{definition}[Permutation relation]
    \label{def:permutation-relation}
    \leavevmode
    A relation $\approx$ is a correct permutation relation iff it:
    \begin{itemize}
        \item is reflexive, symmetric, transitive (an equivalence),
        \item is a congruence wrt $\mult$: $a \mult b \to c \mult d \to a \mult c \approx b \mult d$,
        \item is commutative: $a \mult b \approx b \mult a$, and
        \item respects $\ext{(\blank)}$: $\forall f, \, a \approx b \to \ext{f}(a) = \ext{f}(b)$.
    \end{itemize}
\end{definition}

Let $q : F(A) \to {F(A) / \approx}$ be the quotient map (inclusion into the quotient).
%
The generators map is given by $q \comp \eta_A$, the identity element is $q(e)$,
and the $\doubleplus$ operation can also be lifted to the quotient by congruence.

\begin{proposition}
    $({\str{F}(A) / \approx}, \doubleplus, q(e))$ is a commutative monoid.
\end{proposition}

\begin{proof}
    Since $\approx$ is congruence wrt $\mult$,
    we can lift $\mult : F(A) \to F(A) \to F(A)$ to the quotient to obtain
    $\doubleplus : {F(A) / \approx} \to {F(A) / \approx} \to {F(A) / \approx}$.
    $\doubleplus$ would also satisfy the unit laws and associativity law which $\mult$ satisfy.
    Commutativity of $\doubleplus$ follows from the commutativity requirement of $\approx$,
    therefore $({F(A) / \approx}, \doubleplus, q(i))$ forms a commutative monoid.
\end{proof}

For clarity, we will use $\exthat{(\blank)}$ to denote the extension operation of $F(A)$,
and $\ext{(\blank)}$ for the extension operation of ${F(A) / \approx}$.

\begin{definition}
    Given a commutative monoid $\str{X}$ and a map $f : A \to X$,
    we define
    $\ext{f} : {\str{F}(A) / \approx} \; \to \mathfrak{X}$ as follows:
    we first obtain $\exthat{f} : \str{F}(A) \to \mathfrak{X}$ by universal property of $F$, and lift it
    to ${\str{F}(A)/\approx} \to \mathfrak{X}$, which is allowed since $\approx$ respects $\ext{(\blank)}$.
\end{definition}

Using the correct specification of a permutation relation, we can prove that $({\str{F}(A) / \approx})$ gives the free
commutative monoid on $A$.

\begin{lemma}
    For all $f : A \to X$, $x : F(A)$, $\ext{f}(q(x))$ reduces to $\exthat{f}(x)$.
\end{lemma}

\begin{proposition}[Universal property for ${\str{F}(A) / \approx}$]
    $({\str{F}(A) / \approx},\eta_A : A \xto{q} \str{F}(A) \to {\str{F}(A) / \approx})$
    is the free commutative monoid on $A$.
\end{proposition}

\begin{proof}
    To show that $\ext{(\blank)}$ is an inverse to $\blank \comp \eta_A$,
    we first show $\ext{(\blank)}$ is the right inverse to $\blank \comp \eta_A$.
    For all $f$ and $x$, $(\ext{f} \comp \eta_A)(x) = (\ext{f} \comp q)(\mu_A(x)) = \exthat{f}(\mu_A(x))$.
    By universal property of $F$, $\exthat{f}(\mu_A(x)) = f(x)$, therefore $(\ext{f} \comp \eta_A)(x) = f(x)$.
    By function extensionality, for any $f$, $\ext{f} \circ \eta_A = f$,
    and $(\blank \circ \eta_A) \comp \ext{(\blank)} = id$.

    To show $\ext{(\blank)}$ is the left inverse to $\blank \comp \eta_A$, we need to prove
    for any commutative monoid homomorphism $f : {F(A) / \approx} \to \mathfrak{X}$ and $x : {F(A) / \approx}$,
    $\ext{(f \comp \eta_A)}(x) = f(x)$. To prove this it is suffice to show for all $x : F(A)$,
    $\ext{(f \comp \eta_A)}(q(x)) = f(q(x))$.
    $\ext{(f \comp \eta_A)}(q(x))$ reduces to $\exthat{(f \comp q \comp \mu_A)}(x)$.
    Note that both $f$ and $q$ are homomorphism, therefore $f \comp q$ is a homomorphism. By
    universal property of $F$ we get $\exthat{(f \comp q \comp \mu_A)}(x) = (f \comp q)(x)$,
    therefore $\ext{(f \comp \eta_A)}(q(x)) = f(q(x))$.

    We have now shown that $(\blank) \comp \eta_A$ is an equivalence from
    commutative monoid homomorphisms ${F(A) / \approx} \to \mathfrak{X}$
    to set functions $A \to X$, and its inverse is given by $\ext{(\blank)}$, which maps set
    functions $A \to X$ to commutative monoid homomorphisms ${F(A) / \approx} \to \mathfrak{X}$.
    Therefore, ${F(A) / \approx}$ is indeed the free commutative monoid on $A$.
\end{proof}


\subsection{Lists quotiented by permutation}\label{cmon:plist}

A specific instance of this construction would be $\List$ quotiented with permutation to get commutativity. This
construction is also considered in~\cite{joramConstructiveFinalSemantics2023}, who gave a proof that $\PList$ is
equivalent to the free commutative monoid as a HIT. We give a direct proof of its universal property using our
generalistaion.

The permutation relation on lists is as follows, which swaps any two adjacent elements in the middle of the list.
This is only one example of such a relation, of course, there are many such in the literature.
\begin{definition}[PList]
    \leavevmode
    \begin{code}
data Perm (A : UU) : List A -> List A -> UU where
  perm-refl : forall {xs} -> Perm xs xs
  perm-swap : forall {x y xs ys zs}
           -> Perm (xs ++ x :: y :: ys) zs
           -> Perm (xs ++ y :: x :: ys) zs

PList : UU -> UU
PList A = List A / Perm
    \end{code}
\end{definition}

We have already given a proof of $\List$ being the free monoid in~\cref{mon:lists}.
By~\cref{cmon:qfreemon} it suffices to show $\Perm$ satisfies the axioms of permutation relation
to show $\PList$ is the free commutative monoid.

It is trivial to see how $\Perm$ satisfies reflexivity, symmetry, transitivity.
We can also prove $\Perm$ is congruent wrt to $\doubleplus$ inductively.
Commutativity is proven similarly, like the proof for $\SList$ in~\cref{slist:comm}.

\begin{proposition}\label{plist:sharp-sat}
    For all $f : A \to X$, $x, \, y : A$ and $xs, \, ys : \SList(A)$,
    $\ext{f}(xs\,\doubleplus\,x :: y :: ys) = \ext{f}(xs\,\doubleplus\,y :: x :: ys)$.
\end{proposition}

With this we can prove $\Perm$ respects $\ext{(\blank)}$.

\begin{proof}
    We can prove this by induction on $xs$. For $xs = []$, by homomorphism properties of $\ext{f}$,
    we can prove $\ext{f}(x :: y :: ys) = \ext{f}([ x ]) \mult \ext{f}([ y ]) \mult \ext{f}(ys)$.
    Since the image of $\ext{f}$ is a commutative monoid, we have
    $\ext{f}([ x ]) \mult \ext{f}([ y ]) = \ext{f}([ y ]) \mult \ext{f}([ x ])$, therefore proving
    $\ext{f}(x :: y :: ys) = \ext{f}(y :: x :: ys)$. For $xs = z :: zs$, we can prove
    $\ext{f}((z :: zs)\,\doubleplus\,x :: y :: ys) = \ext{f}([ z ]) \mult \ext{f}(zs\,\doubleplus\,x :: y :: ys)$.
    We can then complete the proof by induction to obtain
    $\ext{f}(zs\,\doubleplus\,x :: y :: ys) = \ext{f}(zs\,\doubleplus\,y :: x :: ys)$ and reassembling
    back to $\ext{f}((z :: zs)\,\doubleplus\,y :: x :: ys)$ by homomorphism properties of $\ext{f}$.
\end{proof}

\subsubsection*{Remark on representation}\label{plist:rep}
With this representation it is very easy to lift functions and properties defined on $\List$
to $\PList$ since $\PList$ is a quotient of $\List$. The inductive nature of $\PList$ makes it
very easy to define algorithms and proofs that are inductive in nature, e.g. defining insertion sort
on $\PList$ is very simple since insertion sort inductively sorts a list, which we can easily do by
pattern matching on $\PList$ since the construction of $\PList$ is definitionally inductive.
This property also makes it such that oftentimes inductively constructed $\PList$ would normalize to the
simplest form of the $\PList$, e.g. $q([ x ]) \doubleplus q([y, z])$ normalizes to $q([x,y,z])$ by
definition, saving the efforts of defining auxillary lemmas to prove their equality.

This inductive nature, however, makes it difficult to define efficient operations on $\PList$. Consider a
divide-and-conquer algorithm such as merge sort, which involves partitioning a $\PList$ of length $n+m$ into
two smaller $\PList$ of length $n$ and length $m$. The inductive nature of $\PList$ makes it such that
we must iterate all $n$ elements before we can make such a partition, thus making $\PList$ unintuitive
to work with when we want to reason with operations that involve arbitrary partitioning.

Also, whenever we define a function on $\PList$ by pattern matching we would also need to show
the function respects $\Perm$, i.e. $\Perm as\,bs \to f(as) = f(bs)$. This can be annoying because
of the many auxiliary variables in the constructor $\term{perm-swap}$, namely $xs$, $ys$, $zs$.
We need to show $f$ would respect a swap in the list anywhere between $xs$ and $ys$, which can
unnecessarily complicate the proof. For example in the inductive step of~\cref{plist:sharp-sat},
$\ext{f}((z :: zs)\,\doubleplus\,x :: y :: ys) = \ext{f}([ z ]) \mult \ext{f}(zs\,\doubleplus\,x :: y :: ys)$,
to actually prove this in Cubical Agda would involve first applying associativity to prove
$(z :: zs)\,\doubleplus\,x :: y :: ys = z :: (zs\,\doubleplus\,x :: y :: ys)$, before we can actually
apply homomorphism properties of $f$. In the final reassembling step, similarly,
we also need to re-apply associativity to go from $z :: (zs\,\doubleplus\,y :: x :: ys)$
to $(z :: zs)\,\doubleplus\,y :: x :: ys$. Also since we are working with an equivalence relation we
defined ($\Perm$) and not the equality type directly, we cannot exploit the many combinators defined
in the standard library for the equality type and often needing to re-define combinators ourselves.

\subsection{Swap-List}\label{cmon:slist}

Informally, quotients generate too many points, then quotient out into equivalence classes by the congruence relation.
%
Alternately, HITs use generators (points) and higher generators (paths) (and higher higher generators and so on\ldots).
%
We can define free commutative monoids using HITs were adjacent swaps generate all symmetries,
for example swap-list taken from \cite{choudhuryFreeCommutativeMonoids2023}, where they have also given a proof of its
universal property. We include this presentation for completeness, which is also a good presentation in some of our
proofs.

\begin{code}
data SList (A : UU) : UU where
  nil : SList A
  _cons_ : A -> SList A -> SList A
  swap : forall x y xs -> x :: y :: xs == y :: x :: xs
  trunc : forall x y -> (p q : x == y) -> p == q
\end{code}
\vspace{1em}

The $\term{trunc}$ constructor is necessary to truncate $\SList$ down to a set,
thereby ignoring any higher paths intrudoced by the $\term{swap}$ constructor.
This is opposed to $\List$, which does not need a $\term{trunc}$ constructor
because it does not have any path constructors, therefore it can be proven that $\List(A)$
is a set assuming $A$ is a set (see formalization).

\begin{definition}[Concatenation]
    We define the concatenation operation $\doubleplus : \SList(A) \to \SList(A) \to \SList(A)$
    recursively, where we also have to consider the (functorial) action on the $\term{swap}$ path using $\term{ap}$.
    \begin{align*}
        [] \doubleplus ys                                 & = ys                                   \\
        (x :: xs) \doubleplus ys                          & = x :: (xs \doubleplus ys)             \\
        \term{ap}_{\doubleplus ys}(\term{swap}(x, y, xs)) & = \term{swap}(x, y, ys \doubleplus xs)
    \end{align*}
\end{definition}

\cite{choudhuryFreeCommutativeMonoids2023} have already given a proof of $(\SList(A), \doubleplus, [])$ satisfying
commutative monoid laws. We explain the proof of $\doubleplus$ satisfying commutativity here.

\begin{lemma}[Head rearrange]\label{slist:cons}
    For all $x : A$, $xs : \SList(A)$, $x :: xs = xs \doubleplus [ x ]$.
\end{lemma}

\begin{proof}
    We can prove this by induction on $xs$.
    For $xs = []$ this is trivial. For $xs = y :: ys$, we have the induction hypothesis $x :: ys = ys \doubleplus [ x ]$.
    By applying $y :: \blank$ on both side and then apply $\term{swap}$, we can complete the proof.
\end{proof}

\begin{theorem}[Commutativity]\label{slist:comm}
    For all $xs,\,ys : \SList(A)$, $xs \doubleplus ys = ys \doubleplus xs$.
\end{theorem}

\begin{proof}
    By induction on $xs$ we can iteratively apply~\cref{slist:cons} to move all elements of $xs$
    to after $ys$. This would move $ys$ to the head and $xs$ to the end, thereby proving
    $xs \doubleplus ys = ys \doubleplus xs$.
\end{proof}

\subsubsection*{Remark on representation}\label{slist:rep}
Much like $\PList$ and $\List$, $\SList$ is inductively defined, therefore making it very intuitive to reason
with when defining inductive operations or inductive proofs on $\SList$, however difficult to reason with
when defining operations that involve arbitrary partitioning, for reasons similar to those given
in~\cref{plist:rep}.

Unlike $\PList$ which is defined as a set quotient, this is defined as a HIT, therefore handling equalities
between $\SList$ is much simpler than $\PList$. We would still need to prove a function $f$ respects
the path constructor of $\SList$ when pattern matching, i.e. $f(x :: y :: xs) = f(y :: x :: xs)$.
Unlike $\PList$ we do not need to worry about as many auxiliary variables since swap
only happens at the head of the list, whereas with $\PList$ we would need to prove
$f(xs\,\doubleplus\,x :: y :: ys) = f(xs\,\doubleplus\,y :: x :: ys)$. One may be tempted to just remove $xs$
from the $\term{perm-swap}$ constructor such that it becomes
$\term{perm-swap} : \forall x\,y\,ys\,zs \to \Perm\,(x :: y :: ys)\,zs \to \Perm\,(y :: x :: ys)\,zs$.
However this would break $\Perm$'s congruence wrt to $\doubleplus$, therefore violating the axioms of
permutation relations. Also, since we are working with the identity type directly, properties we would
expect from $\term{swap}$, such as reflexivity, transitivity, symmetry, congruence and such all comes directly by
construction, whereas with $\Perm$ we would have to prove these properties manually.
We can also use the many combinatorics defined in the standard library for equational reasoning,
making the handling of $\SList$ equalities a lot simpler.

\subsection{Bag}\label{cmon:bag}
Alternatively, we can also quotient $\Array$ with symmetries to get commutativity.
%
This construction is first considered in~\cite{altenkirchDefinableQuotientsType2011}
and~\cite{liQuotientTypesType2015}, then partially considered in~\cite{choudhuryFreeCommutativeMonoids2023},
and also in~\cite{joramConstructiveFinalSemantics2023},
who gave a similar construction, where only the index function is quotiented, instead of
the entire array.
%
\cite{danielssonBagEquivalenceProofRelevant2012} also considered $\Bag$ as a setoid relation
on $\List$ in an intensional MLTT setting.
%
\cite{joramConstructiveFinalSemantics2023} prove that their version of $\Bag$ is the free commutative monoid by
equivalence to the other HIT constructions.
%
We give a direct proof of its universal property instead, using the technology we have developed.

\begin{definition}[Bag]
    \label{def:bag}
    \leavevmode
    \begin{code}
_$\approx$_ : Array A $\to$ Array A $\to$ UU
(n , f) $\approx$ (m , g) = Sg(sig : Fin n eqv Fin m) v = w comp sig

Bag : UU -> UU
Bag A = Array A / _$\approx$_
    \end{code}
\end{definition}

Note that by the pigeonhole principle, $\sigma$ can only be constructed when $n = m$, though this requires a proof in
type theory (see the formalization).
%
Conceptually, we are quotienting array by a permutation or an automorphism on the indices.

We have already given a proof of $\Array$ being the free monoid in~\cref{mon:array}.
%
By~\cref{cmon:qfreemon} it suffices to show $\approx$ satisfies the axioms of permutation relations to show that $\Bag$
is the free commutative monoid.

\begin{proposition}
    $\approx$ is a equivalence relation.
\end{proposition}

\begin{proof}
    We can show any array $xs$ is related to itself by the identity isomorphism, therefore $\approx$ is reflexive.
    If $xs \approx ys$ by $\sigma$, we can show $ys \approx xs$ by $\sigma^{-1}$, therefore $\approx$ is symmetric.
    If $xs \approx ys$ by $\sigma$ and $ys \approx zs$ by $\phi$, we can show $xs \approx zs$ by $\sigma \comp \phi$,
    therefore $\approx$ is transitive.
\end{proof}

\begin{proposition}\label{bag:cong}
    $\approx$ is congruent wrt to $\doubleplus$.
\end{proposition}

\begin{proof}
    Given $(n, f) \approx (m, g)$ by $\sigma$ and $(u, p) \approx (v, q)$ by $\phi$,
    we want to show $(n, f) \doubleplus (u, p) \approx (m, g) \doubleplus (v, q)$ by some $\tau$.
    We can construct $\tau$ as follow:

    \begin{align*}
        \tau := \Fin[n+u] \xto{\sim} \Fin[n] + \Fin[u] \xto{\sigma,\,\phi} \Fin[m] + \Fin[v] \xto{\sim} \Fin[m+v] \\
    \end{align*}

    \begin{figure}[H]
        \centering
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXHswLDEsXFxkb3RzLG4tMSwgbixuKzEsXFxkb3RzLG4rbS0xXFx9Il0sWzAsMSwiIFxce24sbisxXFxkb3RzLG4rbS0xLDAsMSxcXGRvdHMsbi0xXFx9Il0sWzAsMSwiIiwwLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dXQ==
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0,1,\dots,n-1, \color{blue} n,n+1,\dots,n+u-1 \color{black}\}} \\
            { \{\color{red}\sigma(0),\sigma(1)\dots,\sigma(n-1), \color{blue}\phi(0),\phi(1),\dots,\phi(u-1) \color{black}\}}
            \arrow["{\sigma,\phi}", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \caption{Operation of $\tau$}
        \label{fig:enter-label}
    \end{figure}

\end{proof}

\begin{proposition}\label{bag:comm}
    $\approx$ is commutative.
\end{proposition}

\begin{proof}
    We want to show for any arrays $(n, f)$ and $(m, g)$, $(n, f) \mult (m, g) \approx (m, g) \mult (n, f)$
    by some $\phi$.

    We can use combinators from formal operations in symmetric rig groupoids \cite{choudhurySymmetriesReversibleProgramming2022} to define $\phi$:
    \begin{align*}
        \phi := \Fin[n+m] \xto{\sim} \Fin[n] + \Fin[m] \xto{\term{swap}_{+}} \Fin[m] + \Fin[n] \xto{\sim} \Fin[m+n] \\
    \end{align*}

    \begin{figure}[H]
        \centering
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXHswLDEsXFxkb3RzLG4tMSwgbixuKzEsXFxkb3RzLG4rbS0xXFx9Il0sWzAsMSwiIFxce24sbisxXFxkb3RzLG4rbS0xLDAsMSxcXGRvdHMsbi0xXFx9Il0sWzAsMSwiIiwwLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dXQ==
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0,1,\dots,n-1, \color{blue} n,n+1,\dots,n+m-1 \color{black}\}} \\
            { \{\color{blue}n,n+1\dots,n+m-1, \color{red}0,1,\dots,n-1 \color{black}\}}
            \arrow["\phi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \caption{Operation of $\phi$}
        \label{fig:enter-label}
    \end{figure}

\end{proof}

\begin{proposition}
    $\approx$ respects $\ext{(\blank)}$ for arrays.
\end{proposition}

It suffices to show that $\ext{f}$ is invariant under permutation: for all $\phi\colon \Fin[n]\xto{\sim}\Fin[n]$,
$\ext{f}(n, i) \id \ext{f}(n, i \circ \phi)$. To prove this we first need to develop some formal combinatorics of
\emph{punching in} and \emph{punching out} indices. These operations are developed further
in~\cite{choudhurySymmetriesReversibleProgramming2022} for studying permutation codes.

\begin{lemma}\label{bag:tau}
    Given $\phi\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$, there is a $\tau\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$
    such that $\tau(0) = 0$, and $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$.
\end{lemma}

\begin{proof}
    Let $k$ be $\phi^{-1}(0)$, and $k + j = S(n)$, we can construct $\tau$ as follow:
    \begin{align*}
        \tau := \Fin[S(n)] \xto{\phi} \Fin[S(n)] \xto{\sim} \Fin[k+j] \xto{\sim} \Fin[k] + \Fin[j]
        \xto{\term{swap}_{+}} \Fin[j] + \Fin[k] \xto{\sim} \Fin[j+k] \xto{\sim} \Fin[S(n)]
    \end{align*}

    \begin{figure}[H]
        \centering
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, 1, 2, \dots, \color{red}k, k+1, k+2, \dots \color{black}\}} \\
            { \{\color{blue}x, y, z, \dots, \color{red}0, u, v, \dots \color{black}\}}
            \arrow["\phi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \hspace{1em}
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, 1, 2, \dots, \color{red}k, k+1, k+2, \dots \color{black}\}} \\
            { \{\color{red}0, u, v, \dots, \color{blue}x, y, z, \dots \color{black}\}}
            \arrow["\tau", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \caption{Operation of $\tau$ constructed from $\phi$}
        \label{fig:enter-label}
    \end{figure}

    It is trivial to prove $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$ since the only
    operation on indices in $\tau$ is $\term{swap}_{+}$. It suffices to show $(S(n), i \circ \phi)$
    can be decomposed into two arrays such that $(S(n), i \circ \phi) = (k, g) \doubleplus (j, h)$
    for some $g$ and $h$. Since the image of $\ext{f}$ is a commutative monoid and $\ext{f}$ is a homomorphism,
    $\ext{f}((k, g) \doubleplus (j, h)) = \ext{f}(k, g) \mult \ext{f}(j, h) = \ext{f}(j, h) \mult \ext{f}(k, g) =
        \ext{f}((j, h) \doubleplus (k, g))$, thereby proving $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$.

\end{proof}

\begin{lemma}\label{bag:punch}
    Given $\tau\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$ where $\tau(0) = 0$,
    there is a $\psi : \Fin[n] \xto{\sim} \Fin[n]$ such that $\tau \circ S = S \circ \psi$.
\end{lemma}

\begin{proof}
    We can construct $\psi$ by $\psi(x) = \tau(S(x)) - 1$.
    %
    Since $\tau$ maps only 0 to 0 by assumption, $\forall x. \, \tau(S(x)) > 0$, therefore
    the $(- 1)$ is well defined. This is the special case for $k = 0$ in the punch-in and punch-out
    equivalence for Lehmer codes in~\cite{choudhurySymmetriesReversibleProgramming2022}.

    \begin{figure}[H]
        \centering
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, \color{red}1, 2, 3, \dots \color{black}\}} \\
            { \{\color{blue}0, \color{red} x, y, z \dots \color{black}\}}
            \arrow["\tau", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \hspace{1em}
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0, 1, 2, \dots \color{black}\}} \\
            { \{\color{red} x-1, y-1, z-1 \dots \color{black}\}}
            \arrow["\psi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \caption{Operation of $\psi$ constructed from $\tau$}
        \label{fig:enter-label}
    \end{figure}
\end{proof}

We now prove our main theorem.

\begin{theorem}[Permutation invariance]\label{bag:perm-sat}
    For all $\phi\colon \Fin[n]\xto{\sim}\Fin[n]$, $\ext{f}(n, i) \id \ext{f}(n, i \circ \phi)$.
\end{theorem}

\begin{proof}
    We can prove this by induction on $n$.
    \begin{itemize}
        \item On $n = 0$, $\ext{f}(0, i) \id \ext{f}(0, i \circ \phi) = e$.
        \item On $n = S(m)$,
              \begin{align*}
                   & \ext{f}(S(m), i \circ \phi)                                                                       \\
                   & = \ext{f}(S(m), i \circ \tau)                          & \text{by~\cref{bag:tau}}                 \\
                   & = f(i(\tau(0))) \mult \ext{f}(m, i \circ \tau \circ S) & \text{by definition of $\ext{(\blank)}$} \\
                   & = f(i(0)) \mult \ext{f}(m, i \circ \tau \circ S)       & \text{by construction of $\tau$}         \\
                   & = f(i(0)) \mult \ext{f}(m, i \circ S \circ \psi)       & \text{by~\cref{bag:punch}}               \\
                   & = f(i(0)) \mult \ext{f}(m, i \circ S)                  & \text{induction}                         \\
                   & = \ext{f}(S(m), i)                                     & \text{by definition of $\ext{(\blank)}$}
              \end{align*}
    \end{itemize}
\end{proof}

\subsubsection*{Remark on representation}\label{bag:rep}

Unlike $\PList$ and $\SList$, $\Bag$ and its underlying construction $\Array$ are not inductively defined,
making it difficult to work with when trying to do induction on them. For example,
in the proof~\cref{array:univ}, two lemmas~\cref{array:eta-suc} and~\cref{array:split} are needed to do
induction on $\Array$, as opposed to $\List$ and its quotients, where we can do induction simply by
pattern matching. Much like $\PList$, when defining functions on $\Bag$, we need to show they respect
$\approx$, i.e. $as \approx bs \to f(as) = f(bs)$. This is notably much more difficult than
$\PList$ and $\SList$, because whereas with $\PList$ and $\SList$ we only need to consider "small permutations"
(i.e. swapping adjacent elements), with $\Bag$ we need to consider all possible permutations. For example,
in the proof of~\cref{bag:perm-sat}, we need to first construct a $\tau$ which satisfies $\tau(0) = 0$ and prove
$\ext{f}(n, i \comp \sigma) = \ext{f}(n, i \comp \tau)$ before we can apply induction.

Also, on a more technical note, since $\Array$ and $\Bag$ are not simple data types, the definition of
the monoid operation on them $\doubleplus$ are necessarily more complicated, and unlike $\List$, $\PList$
and $\SList$, constructions of $\Array$ and $\Bag$ via $\doubleplus$ often would not normalize into a
very simple form, but would instead expand into giant trees of terms. This makes it such that when working
with $\Array$ and $\Bag$ we need to be very careful or otherwise Agda would be stuck trying to display
the normalized form of $\Array$ and $\Bag$ in the goal and context menu. Type-checking also becomes a lengthy
process that tests if the user possesses the virtue of patience.

However, performing arbitrary partitioning with $\Array$ and $\Bag$ is much easier than
$\List$, $\SList$, $\PList$. For example,
one can simply use the combinator $\Fin[n+m] \xto{\sim} \Fin[n] + \Fin[m]$ to partition the array,
then perform operations on the partitions such as swapping in~\cref{bag:comm},
or perform operations on the partitions individually such as two individual permutation in~\cref{bag:cong}.
This makes it such that when defining divide-and-conquer algorithms such as merge sort,
$\Bag$ and $\Array$ are more natural to work with than $\List$, $\SList$, and $\PList$.
