% !TEX root = cpp25-sort.tex

\section{Constructions of Free Comm. Monoids}
\label{sec:commutative-monoids}

The next step is to add commutativity to each construction of free monoids.
%
Informally, adding commutativity to free monoids turns ``ordered lists'' to ``unordered lists'',
where the ordering is the one imposed by the position or index of the elements in the list.
%
This is crucial to our goal of studying sorting,
as we will study sorting as a function mapping back unordered lists to ordered lists,
which is later in~\cref{sec:sorting}.

It is known that finite multisets are (free) commutative monoids,
under the operation of multiset union: $xs \cup ys = ys \cup xs$.
%
The order is ``forgotten'' in the sense that it doesn't matter how two multisets are union-ed together,
such as, $\bag{a, a, b, c} = \bag{b, a, c, a}$ are equal as finite multisets
(justifying the bag notation).
%
This is unlike free monoids,
where $[a, a, b, c] \neq [b, a, c, a]$ (justifying the list notation).

\subsection{Free monoids with a quotient}\label{cmon:qfreemon}

Instead of constructing free commutative monoids directly,
the first construction we study is to take \emph{any} free monoid and quotient by \emph{symmetries}.
%
Specific instances of this construction are given in~\cref{cmon:plist,cmon:bag}.

From the universal algebraic perspective developed in~\cref{sec:universal-algebra},
this means to extend the equational theory of a given algebraic signature with symmetries.
%
If $(\str{F}(A), \eta)$ is a free monoid construction satisfying its universal property,
then we'd like to quotient $F(A)$ by an \emph{appropriate} symmetry relation $\approx$.
%
This is exactly a \emph{permutation relation}!

% $(\str{F}(A), i, \mult)$ quotiented by a permutation relation $({F(A) / \approx}, e, \concat)$
% would be the free commutative monoid on $A$.

\begin{definition}[Permutation relation]
    \label{def:permutation-relation}
    \leavevmode
    A binary relation $\approx$ on free monoids is a correct permutation relation iff it:
    \begin{itemize}
        \item is reflexive, symmetric, transitive (an equivalence),
        \item is a congruence wrt $\mult$: $a \approx b \to c \approx d \to a \mult c \approx b \mult d$,
        \item is commutative: $a \mult b \approx b \mult a$, and
        \item respects $\ext{(\blank)}$: $\forall f, \, a \approx b \to \ext{f}(a) = \ext{f}(b)$.
    \end{itemize}
\end{definition}

Let $q : F(A) \twoheadrightarrow \quot{F(A)}{\approx}$ be the quotient (inclusion) map.
%
The generators map is given by $q \comp \eta_A$, the identity element is $q(e)$,
and the $\mult$ operation is lifted to the quotient by congruence.

\begin{propositionrep}
    $(\quot{\str{F}(A)}{\approx}, \mult, q(e))$ is a commutative monoid.
\end{propositionrep}
\begin{proof}
    Since $\approx$ is a congruence wrt $\mult$,
    we can lift $\mult : F(A) \to F(A) \to F(A)$ to the quotient to obtain
    $\concat : \quot{F(A)}{\approx} \to \quot{F(A)}{\approx} \to \quot{F(A)}{\approx}$.
    $\concat$ also satisfies the unit and associativity laws that $\mult$ satisfy.
    Commutativity of $\concat$ follows from the commutativity requirement of $\approx$,
    therefore $(\quot{F(A)}{\approx}, \concat, q(i))$ forms a commutative monoid.
\end{proof}
\noindent
For clarity, we will use $\exthat{(\blank)}$ to denote the extension operation of $F(A)$,
and $\ext{(\blank)}$ for the extension operation of $\quot{F(A)}{\approx}$.

\begin{definition}
    Given a commutative monoid $\str{X}$ and a map $f : A \to X$,
    we define
    $\ext{f} : \quot{\str{F}(A)}{\approx} \; \to \str{X}$ as follows:
    we first obtain $\exthat{f} : \str{F}(A) \to \str{X}$ by universal property of $F$, and lift it
    to $\quot{\str{F}(A)}{\approx} \to \str{X}$, which is allowed since $\approx$ respects $\ext{(\blank)}$.
\end{definition}

\begin{propositionrep}[Universal property for $\quot{\str{F}(A)}{\approx}$]
    \label{prop:qfreemon}
    \leavevmode
    $(\quot{\str{F}(A)}{\approx}, \eta_A : {A \xto{\eta_A} \str{F}(A) \xto{q} \quot{\str{F}(A)}{\approx}})$
    is the free comm. monoid on $A$.
\end{propositionrep}
\begin{proof}
    To show that $\ext{(\blank)}$ is an inverse to $(\blank) \comp \eta_A$,
    we first show $\ext{(\blank)}$ is the right inverse to $(\blank) \comp \eta_A$.
    For all $f$ and $x$, $(\ext{f} \comp \eta_A)(x) = (\ext{f} \comp q)(\mu_A(x)) = \exthat{f}(\mu_A(x))$.
    By universal property of $F$, $\exthat{f}(\mu_A(x)) = f(x)$, therefore $(\ext{f} \comp \eta_A)(x) = f(x)$.
    By function extensionality, for any $f$, $\ext{f} \circ \eta_A = f$,
    and $(\blank \circ \eta_A) \comp \ext{(\blank)} = id$.

    To show $\ext{(\blank)}$ is the left inverse to $(\blank) \comp \eta_A$, we need to prove
    for any commutative monoid homomorphism $f : {\str{F}(A)}{\approx} \to \str{X}$ and $x : {\str{F}(A)}{\approx}$,
    $\ext{(f \comp \eta_A)}(x) = f(x)$. To prove this it is suffice to show for all $x : \str{F}(A)$,
    $\ext{(f \comp \eta_A)}(q(x)) = f(q(x))$.
    $\ext{(f \comp \eta_A)}(q(x))$ reduces to $\exthat{(f \comp q \comp \mu_A)}(x)$.
    Note that both $f$ and $q$ are homomorphism, therefore $f \comp q$ is a homomorphism. By
    universal property of $F$ we get $\exthat{(f \comp q \comp \mu_A)}(x) = (f \comp q)(x)$,
    therefore $\ext{(f \comp \eta_A)}(q(x)) = f(q(x))$.

    We have now shown that $(\blank) \comp \eta_A$ is an equivalence from
    commutative monoid homomorphisms $\quot{\str{F}(A)}{\approx} \to \str{X}$
    to set functions $A \to X$, and its inverse is given by $\ext{(\blank)}$, which maps set
    functions $A \to X$ to commutative monoid homomorphisms $\quot{\str{F}(A)}{\approx} \to \str{X}$.
    Therefore, $\quot{\str{F}(A)}{\approx}$ is indeed the free commutative monoid on $A$.
\end{proof}


\subsection{Lists quotiented by permutation}\label{cmon:plist}

A specific instance of this construction is $\List$ quotiented by a permutation relation to get commutativity.
We study one such construction ($\PList$), considered in~\cite{joramConstructiveFinalSemantics2023},
who give a proof that $\PList$ is equivalent to the free commutative monoid (constructed as a HIT).
We give a direct proof of its universal property using our generalisation.

Of course, there are many permutation relations in the literature, we consider
the one which swaps any two adjacent elements somewhere in the middle of the list.
\begin{definition}[PList]
    \leavevmode
    \begin{code}
data Perm (A : UU) : List A -> List A -> UU where
  perm-refl : forall {xs} -> Perm xs xs
  perm-swap : forall {x y xs ys zs}
           -> Perm (xs concat x cons y cons ys) zs
           -> Perm (xs concat y cons x cons ys) zs

PList : UU -> UU
PList A = List A quot Perm
    \end{code}
\end{definition}
\vspace*{-2em}
By~\cref{cmon:qfreemon}, it suffices to show $\Perm$ satisfies the axioms of permutation relation
to show $\PList$ is the free commutative monoid.

\begin{propositionrep}
    \label{plist:sharp-sat}
    Let $\str{X}$ be a commutative monoid, and $f : A \to X$.
    For $x,y : A$ and $xs,ys : \PList(A)$,
    $\ext{f}(xs\,\concat\,x \cons y \cons ys) \id \ext{f}(xs\,\concat\,y \cons x \cons ys)$.
    Hence, $\Perm$ respects $\ext{(\blank)}$.
\end{propositionrep}
\begin{proof}
    We can prove this by induction on $xs$. For $xs = []$, by homomorphism properties of $\ext{f}$,
    we can prove $\ext{f}(x \cons y \cons ys) = \ext{f}([ x ]) \mult \ext{f}([ y ]) \mult \ext{f}(ys)$.
    Since the image of $\ext{f}$ is a commutative monoid, we have
    $\ext{f}([ x ]) \mult \ext{f}([ y ]) = \ext{f}([ y ]) \mult \ext{f}([ x ])$, therefore proving
    $\ext{f}(x \cons y \cons ys) = \ext{f}(y \cons x \cons ys)$. For $xs = z \cons zs$, we can prove
    $\ext{f}((z \cons zs)\,\concat\,x \cons y \cons ys) = \ext{f}([ z ]) \mult \ext{f}(zs\,\concat\,x \cons y \cons ys)$.
    We can then complete the proof by induction to obtain
    $\ext{f}(zs\,\concat\,x \cons y \cons ys) = \ext{f}(zs\,\concat\,y \cons x \cons ys)$ and reassembling
    back to $\ext{f}((z \cons zs)\,\concat\,y \cons x \cons ys)$ by homomorphism properties of $\ext{f}$.
\end{proof}

\subsubsection*{Remarks}\label{plist:rep}
With this representation it is very easy to lift functions and properties defined on $\List$
to $\PList$ since $\PList$ is a quotient of $\List$. The inductive nature of $\PList$ makes it
very easy to define algorithms and proofs that are inductive in nature, e.g. defining insertion sort
on $\PList$ is very simple since insertion sort inductively sorts a list, which we can easily do by
pattern matching on $\PList$ since the construction of $\PList$ is definitionally inductive.
This property also makes it such that oftentimes inductively constructed $\PList$ would normalize to the
simplest form of the $\PList$, e.g. $q([ x ]) \concat q([y, z])$ normalizes to $q([x,y,z])$ by
definition, saving the efforts of defining auxillary lemmas to prove their equality.

This inductive nature, however, makes it difficult to define efficient operations on $\PList$. Consider a
divide-and-conquer algorithm such as merge sort, which involves partitioning a $\PList$ of length $n+m$ into
two smaller $\PList$ of length $n$ and length $m$. The inductive nature of $\PList$ makes it such that
we must iterate all $n$ elements before we can make such a partition, thus making $\PList$ unintuitive
to work with when we want to reason with operations that involve arbitrary partitioning.
\begin{toappendix}
    Also, whenever we define a function on $\PList$ by pattern matching we would also need to show
    the function respects $\Perm$, i.e. $\Perm as\,bs \to f(as) = f(bs)$. This can be annoying because
    of the many auxiliary variables in the constructor $\term{perm-swap}$, namely $xs$, $ys$, $zs$.
    We need to show $f$ would respect a swap in the list anywhere between $xs$ and $ys$, which can
    unnecessarily complicate the proof. For example in the inductive step of~\cref{plist:sharp-sat},
    $\ext{f}((z \cons zs)\,\concat\,x \cons y \cons ys) = \ext{f}([ z ]) \mult \ext{f}(zs\,\concat\,x \cons y \cons ys)$,
    to actually prove this in Cubical Agda would involve first applying associativity to prove
    $(z \cons zs)\,\concat\,x \cons y \cons ys = z \cons (zs\,\concat\,x \cons y \cons ys)$, before we can actually
    apply homomorphism properties of $f$. In the final reassembling step, similarly,
    we also need to re-apply associativity to go from $z \cons (zs\,\concat\,y \cons x \cons ys)$
    to $(z \cons zs)\,\concat\,y \cons x \cons ys$. Also since we are working with an equivalence relation we
    defined ($\Perm$) and not the equality type directly, we cannot exploit the many combinators defined
    in the standard library for the equality type and often needing to re-define combinators ourselves.
\end{toappendix}

\subsection{Swap-List}\label{cmon:slist}

Informally, quotients are defined by generating all the points, then quotienting out into equivalence classes by the
congruence relation.
%
Alternately, HITs use generators (points) and higher generators (paths) (and higher higher generators and so on\ldots).
%
We can define free commutative monoids using HITs were adjacent swaps generate all symmetries,
for example swap-lists taken from \cite{choudhuryFreeCommutativeMonoids2023} (and in the Cubical library).
\begin{code}
data SList (A : UU) : UU where
  nil : SList A
  _cons_ : A -> SList A -> SList A
  swap : forall x y xs -> x cons y cons xs == y cons x cons xs
  trunc : forall x y -> (p q : x == y) -> p == q
\end{code}

\begin{toappendix}
    The $\term{trunc}$ constructor is necessary to truncate $\SList$ down to a set,
    thereby ignoring any higher paths introduced by the $\term{swap}$ constructor.
    This is opposed to $\List$, which does not need a $\term{trunc}$ constructor
    because it does not have any path constructors, therefore it can be proven that $\List(A)$
    is a set assuming $A$ is a set (see formalization).

    \begin{definition}[Concatenation]
        We define the concatenation operation $\concat : \SList(A) \to \SList(A) \to \SList(A)$
        recursively, where we also have to consider the (functorial) action on the $\term{swap}$ path using $\term{ap}$.
        \begin{align*}
            [] \concat ys                                 & = ys                               \\
            (x \cons xs) \concat ys                       & = x \cons (xs \concat ys)          \\
            \term{ap}_{\concat ys}(\term{swap}(x, y, xs)) & = \term{swap}(x, y, ys \concat xs)
        \end{align*}
    \end{definition}

    \cite{choudhuryFreeCommutativeMonoids2023} have already given a proof of $(\SList(A), \concat, [])$ satisfying
    commutative monoid laws. We explain the proof of $\concat$ satisfying commutativity here.

    \begin{lemma}[Head rearrange]\label{slist:cons}
        For all $x : A$, $xs : \SList(A)$, $x \cons xs = xs \concat [ x ]$.
    \end{lemma}

    \begin{proof}
        We can prove this by induction on $xs$.
        For $xs \jdgeq []$ this is trivial. For $xs \jdgeq y \cons ys$, we have the induction hypothesis $x \cons ys = ys \concat [ x ]$.
        By applying $y \cons (\blank)$ on both side and then apply $\term{swap}$, we can complete the proof.
    \end{proof}

    \begin{theorem}[Commutativity]\label{slist:comm}
        For all $xs,\,ys : \SList(A)$, $xs \concat ys = ys \concat xs$.
    \end{theorem}

    \begin{proof}
        By induction on $xs$ we can iteratively apply~\cref{slist:cons} to move all elements of $xs$
        to after $ys$. This would move $ys$ to the head and $xs$ to the end, thereby proving
        $xs \concat ys = ys \concat xs$.
    \end{proof}
\end{toappendix}

\subsubsection*{Remarks}\label{slist:rep}
Much like $\PList$ and $\List$, $\SList$ is inductively defined, therefore making it very intuitive to reason
with when defining inductive operations or inductive proofs on $\SList$, however difficult to reason with
when defining operations that involve arbitrary partitioning, for reasons similar to those given
in~\cref{plist:rep}.
\begin{toappendix}
    Unlike $\PList$ which is defined as a set quotient, this is defined as a HIT, therefore handling equalities
    between $\SList$ is much simpler than $\PList$. We would still need to prove a function $f$ respects
    the path constructor of $\SList$ when pattern matching, i.e. $f(x \cons y \cons xs) = f(y \cons x \cons xs)$.
    Unlike $\PList$ we do not need to worry about as many auxiliary variables since swap
    only happens at the head of the list, whereas with $\PList$ we would need to prove
    $f(xs\,\concat\,x \cons y \cons ys) = f(xs\,\concat\,y \cons x \cons ys)$. One may be tempted to just remove $xs$
    from the $\term{perm-swap}$ constructor such that it becomes
    $\term{perm-swap} : \forall x\,y\,ys\,zs \to \Perm\,(x \cons y \cons ys)\,zs \to \Perm\,(y \cons x \cons ys)\,zs$.
    However this would break $\Perm$'s congruence wrt to $\concat$, therefore violating the axioms of
    permutation relations. Also, since we are working with the identity type directly, properties we would
    expect from $\term{swap}$, such as reflexivity, transitivity, symmetry, congruence and such all comes directly by
    construction, whereas with $\Perm$ we would have to prove these properties manually.
    We can also use the many combinatorics defined in the standard library for equational reasoning,
    making the handling of $\SList$ equalities a lot simpler.
\end{toappendix}

\subsection{Bag}\label{cmon:bag}
Alternatively, we can also quotient $\Array$ by symmetries to get commutativity.
%
This construction is first considered in~\cite{altenkirchDefinableQuotientsType2011}
and~\cite{liQuotientTypesType2015}, then partially considered in~\cite{choudhuryFreeCommutativeMonoids2023},
and also in~\cite{joramConstructiveFinalSemantics2023},
who gave a similar construction, where only the index function is quotiented, instead of
the entire array.
%
\cite{danielssonBagEquivalenceProofRelevant2012} also considered $\Bag$ as a setoid relation
on $\List$ in an intensional MLTT setting.
%
\cite{joramConstructiveFinalSemantics2023} prove that their version of $\Bag$
is the free commutative monoid by equivalence to the other HIT constructions.
%
We give a direct proof of its universal property instead, using the technology we have developed.

\begin{definition}[Bag]
    \label{def:bag}
    \leavevmode
    \begin{code}
_$\approx$_ : Array A $\to$ Array A $\to$ UU
(n , f) $\approx$ (m , g) = Sg(sig : Fin n eqv Fin m) v = w comp sig

Bag : UU -> UU
Bag A = Array A quot _$\approx$_
    \end{code}
\end{definition}
\vspace*{-2em}
Note that by the pigeonhole principle, $\sigma$ can only be constructed when $n = m$, though this requires a proof in
type theory (see the formalization).
%
Conceptually, we are quotienting $\Array$ by an automorphism on the indices.

We have already given a proof of $\Array$ being the free monoid in~\cref{mon:array}.
%
By~\cref{cmon:qfreemon} it suffices to show $\approx$ satisfies the axioms of permutation relations to show that $\Bag$
is the free commutative monoid.

\begin{propositionrep}
    $\approx$ is a equivalence relation.
\end{propositionrep}
\begin{proof}
    We can show any array $xs$ is related to itself by the identity isomorphism, therefore $\approx$ is reflexive.
    If $xs \approx ys$ by $\sigma$, we can show $ys \approx xs$ by $\sigma^{-1}$, therefore $\approx$ is symmetric.
    If $xs \approx ys$ by $\sigma$ and $ys \approx zs$ by $\phi$, we can show $xs \approx zs$ by $\sigma \comp \phi$,
    therefore $\approx$ is transitive.
\end{proof}

\begin{proposition}\label{bag:cong}
    $\approx$ is congruent wrt to $\concat$.
\end{proposition}

\begin{proof}
    Given $(n, f) \approx (m, g)$ by $\sigma$ and $(u, p) \approx (v, q)$ by $\phi$,
    we want to show $(n, f) \concat (u, p) \approx (m, g) \concat (v, q)$ by some $\tau$.
    We construct $\tau$ as follows:
    \[
        \tau \defeq \Fin[n+u] \xto{\sim} \Fin[n] + \Fin[u] \xto{\sigma,\,\phi} \Fin[m] + \Fin[v] \xto{\sim} \Fin[m+v] \\
    \]
    which operationally performs:
    \[
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXHswLDEsXFxkb3RzLG4tMSwgbixuKzEsXFxkb3RzLG4rbS0xXFx9Il0sWzAsMSwiIFxce24sbisxXFxkb3RzLG4rbS0xLDAsMSxcXGRvdHMsbi0xXFx9Il0sWzAsMSwiIiwwLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dXQ==
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0,1,\dots,n-1, \color{blue} n,n+1,\dots,n+u-1 \color{black}\}} \\
            { \{\color{red}\sigma(0),\sigma(1)\dots,\sigma(n-1), \color{blue}\phi(0),\phi(1),\dots,\phi(u-1) \color{black}\}}
            \arrow["{\sigma,\phi}", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \enspace.
    \]
\end{proof}

\begin{proposition}\label{bag:comm}
    $\approx$ is commutative.
\end{proposition}

\begin{proof}
    We want to show for any arrays $(n, f)$ and $(m, g)$, $(n, f) \mult (m, g) \approx (m, g) \mult (n, f)$
    by some $\phi$.
    We can use combinators from formal operations in symmetric rig groupoids \cite{choudhurySymmetriesReversibleProgramming2022} to define $\phi$:
    \[
        \phi \defeq \Fin[n+m] \xto{\sim} \Fin[n] + \Fin[m] \xto{\term{swap}_{+}} \Fin[m] + \Fin[n] \xto{\sim} \Fin[m+n] \\
    \]
    which operationally performs:
    \[
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXHswLDEsXFxkb3RzLG4tMSwgbixuKzEsXFxkb3RzLG4rbS0xXFx9Il0sWzAsMSwiIFxce24sbisxXFxkb3RzLG4rbS0xLDAsMSxcXGRvdHMsbi0xXFx9Il0sWzAsMSwiIiwwLHsic3R5bGUiOnsidGFpbCI6eyJuYW1lIjoibWFwcyB0byJ9fX1dXQ==
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0,1,\dots,n-1, \color{blue} n,n+1,\dots,n+m-1 \color{black}\}} \\
            { \{\color{blue}n,n+1\dots,n+m-1, \color{red}0,1,\dots,n-1 \color{black}\}}
            \arrow["\phi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
    \]
\end{proof}

\begin{proposition}
    $\approx$ respects $\ext{(\blank)}$ for arrays.
\end{proposition}
\noindent
It suffices to show that $\ext{f}$ is invariant under permutation: for all $\phi\colon \Fin[n]\xto{\sim}\Fin[n]$,
$\ext{f}(n, i) \id \ext{f}(n, i \circ \phi)$. To prove this we first need to develop some formal combinatorics of
\emph{punching in} and \emph{punching out} indices. These operations are
borrowed from~\cite{mozlerCubicalAgdaSimple2021} and
developed further
in~\cite{choudhurySymmetriesReversibleProgramming2022} for studying permutation codes.

\begin{lemma}\label{bag:tau}
    Given $\phi\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$, there is a permutation $\tau\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$
    such that $\tau(0) = 0$, and $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$.
\end{lemma}

\begin{proof}
    Let $k$ be $\phi^{-1}(0)$, and $k + j = S(n)$, we construct $\tau$:
    \[
        \tau \defeq \Fin[S(n)] \xto{\phi} \Fin[S(n)] \xto{\sim} \Fin[k+j] \xto{\sim} \Fin[k] + \Fin[j]
    \]
    \[
        \xto{\term{swap}_{+}} \Fin[j] + \Fin[k] \xto{\sim} \Fin[j+k] \xto{\sim} \Fin[S(n)]
    \]
    \[
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, 1, 2, \dots, \color{red}k, k+1, k+2, \dots \color{black}\}} \\
            {\{\color{blue}x, y, z, \dots, \color{red}0, u, v, \dots \color{black}\}}
            \arrow["\phi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \\
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, 1, 2, \dots, \color{red}k, k+1, k+2, \dots \color{black}\}} \\
            {\{\color{red}0, u, v, \dots, \color{blue}x, y, z, \dots \color{black}\}}
            \arrow["\tau", maps to, from=1-1, to=2-1]
        \end{tikzcd}
    \]
    It is trivial to show $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$, since the only
    operation on indices in $\tau$ is $\term{swap}_{+}$. It suffices to show $(S(n), i \circ \phi)$
    can be decomposed into two arrays such that $(S(n), i \circ \phi) = (k, g) \concat (j, h)$
    for some $g$ and $h$. Since the image of $\ext{f}$ is a commutative monoid, and $\ext{f}$ is a homomorphism,
    $\ext{f}((k, g) \concat (j, h)) = \ext{f}(k, g) \mult \ext{f}(j, h) = \ext{f}(j, h) \mult \ext{f}(k, g) =
        \ext{f}((j, h) \concat (k, g))$, thereby proving $\ext{f}(S(n), i \circ \phi) = \ext{f}(S(n), i \circ \tau)$.

\end{proof}

\begin{lemma}\label{bag:punch}
    Given $\tau\colon \Fin[S(n)]\xto{\sim}\Fin[S(n)]$ where $\tau(0) = 0$,
    there is a $\psi : \Fin[n] \xto{\sim} \Fin[n]$ such that $\tau \circ S = S \circ \psi$.
\end{lemma}

\begin{proof}
    We construct $\psi$ as $\psi(x) = \tau(S(x)) - 1$.
    %
    Since $\tau$ maps only 0 to 0 by assumption, $\forall x. \, \tau(S(x)) > 0$, therefore
    the $(- 1)$ is well defined. This is the special case for $k = 0$ in the punch-in and punch-out
    equivalence for Lehmer codes in~\cite{choudhurySymmetriesReversibleProgramming2022}.
    \[
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{blue}0, \color{red}1, 2, 3, \dots \color{black}\}} \\
            { \{\color{blue}0, \color{red} x, y, z \dots \color{black}\}}
            \arrow["\tau", maps to, from=1-1, to=2-1]
        \end{tikzcd}
        \\
        \begin{tikzcd}[ampersand replacement=\&,cramped]
            {\{\color{red}0, 1, 2, \dots \color{black}\}} \\
            { \{\color{red} x-1, y-1, z-1 \dots \color{black}\}}
            \arrow["\psi", maps to, from=1-1, to=2-1]
        \end{tikzcd}
    \]
\end{proof}

\begin{theorem}[Permutation invariance]\label{bag:perm-sat}
    For all $\phi\colon \Fin[n]\xto{\sim}\Fin[n]$, $\ext{f}(n, i) \id \ext{f}(n, i \circ \phi)$.
\end{theorem}

\begin{proof}
    By induction on $n$.
    \begin{itemize}
        \item On $n = 0$, $\ext{f}(0, i) \id \ext{f}(0, i \circ \phi) = e$.
        \item On $n = S(m)$,
              \[
                  \begin{array}{rll}
                        & \ext{f}(S(m), i \circ \phi)                          &                                          \\
                      = & \ext{f}(S(m), i \circ \tau)                          & \text{by~\cref{bag:tau}}                 \\
                      = & f(i(\tau(0))) \mult \ext{f}(m, i \circ \tau \circ S) & \text{by definition of $\ext{(\blank)}$} \\
                      = & f(i(0)) \mult \ext{f}(m, i \circ \tau \circ S)       & \text{by construction of $\tau$}         \\
                      = & f(i(0)) \mult \ext{f}(m, i \circ S \circ \psi)       & \text{by~\cref{bag:punch}}               \\
                      = & f(i(0)) \mult \ext{f}(m, i \circ S)                  & \text{induction}                         \\
                      = & \ext{f}(S(m), i)                                     & \text{by definition of $\ext{(\blank)}$}
                  \end{array}
              \]
    \end{itemize}
\end{proof}

\subsubsection*{Remarks}\label{bag:rep}

Unlike $\PList$ and $\SList$, $\Bag$ and its underlying construction $\Array$ are not inductively defined,
making it difficult to work with when trying to do induction on them. For example,
in the proof~\cref{array:univ}, two~\cref{array:eta-suc,array:split} are needed to do
induction on $\Array$, as opposed to $\List$ and its quotients, where we can do induction simply by
pattern matching. Much like $\PList$, when defining functions on $\Bag$, we need to show they respect
$\approx$, i.e. $as \approx bs \to f(as) = f(bs)$. This is notably much more difficult than
$\PList$ and $\SList$, because whereas with $\PList$ and $\SList$ we only need to consider "small permutations"
(i.e. swapping adjacent elements), with $\Bag$ we need to consider all possible permutations. For example,
in the proof of~\cref{bag:perm-sat}, we need to first construct a $\tau$ which satisfies $\tau(0) = 0$ and prove
$\ext{f}(n, i \comp \sigma) = \ext{f}(n, i \comp \tau)$ before we can apply induction.

\begin{toappendix}
    On a more technical note, since $\Array$ and $\Bag$ are not simple data types, the definition of
    the monoid operation on them $\concat$ are necessarily more complicated, and unlike $\List$, $\PList$
    and $\SList$, constructions of $\Array$ and $\Bag$ via $\concat$ often would not normalize into a
    very simple form, but would instead expand into giant trees of terms. This makes it such that when working
    with $\Array$ and $\Bag$ we need to be very careful or otherwise Agda would be stuck trying to display
    the normalized form of $\Array$ and $\Bag$ in the goal and context menu. Type-checking also becomes a lengthy
    process that tests if the user possesses the virtue of patience.

    However, performing arbitrary partitioning with $\Array$ and $\Bag$ is much easier than
    $\List$, $\SList$, $\PList$. For example,
    one can simply use the combinator ${\Fin[n+m] \xto{\sim} \Fin[n] + \Fin[m]}$ to partition the array,
    then perform operations on the partitions such as swapping in~\cref{bag:comm},
    or perform operations on the partitions individually such as two individual permutation in~\cref{bag:cong}.
    This makes it such that when defining divide-and-conquer algorithms such as merge sort,
    $\Bag$ and $\Array$ are more natural to work with than $\List$, $\SList$, and $\PList$.

    We use $\bag{x,y,\dots}$ to denote $\eta_A(x) \mult \eta_A(y) \mult \dots : \MM(A)$,
    and $[x, y, \dots]$ to denote $\eta_A(x) \mult \eta_A(y) \mult \dots : \LL(A)$,
    or $x \cons xs$ to denote $\eta_A(x) \mult xs : \LL(A)$.
\end{toappendix}
